Sparse Function-space Representation \\ of Neural Networks

### 2023-05-17 paper/main.tex
Bold for within-GP test

Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior for the weight space, $q(\vw)$. \todo{This is the first time q(w) is used, should define it}. 
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space by \todo{To be super clear even when reading this part out of context: could specify that using a GP with this mean and covariance},
\begin{equation} 
\label{eq-laplace-approx-function-space} 
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'}, 
\end{equation}
where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}). In \cref{eq-laplace-approx-function-space}, we restrict the function output to a single dimension for notational simplicity, yet this formulation can be readily extended to the multi-output scenario. With this kernel function, we can combine it with the data set $\dataset$ to construct the posterior. \todo{what is "it" here?}
%
\citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the neural network to form a Bayesian GLM model. Both approaches attempted to adjust the GP posterior mean function to ensure the predictions are from the neural network $f_{\vw^*}(\vx)$. 
Contrarily, our methodology intends to make predictions directly from the derived GP, thus evading such adjustments. Next, we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

@@

    % THE TABLE CAN BE FORMATTED MORE LIKE THIS
    % \begin{tabular}{l l C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    % \toprule
    % & Method & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & OOD-AUC~$\uparrow$  \\
    % \midrule
    % \multirow{2}{*}{FMNIST}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \midrule
    % \multirow{2}{*}{CIFAR-10}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \bottomrule
    % \end{tabular}
  \newcommand{\myline}{\protect\tikz[baseline=-.5ex,line width=.7pt]\protect\draw[draw=gray](0,0)--(7em,0);}
  \begin{tabular}{l c C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    \toprule
            &     &  \multicolumn{3}{c}{\myline~MNIST~\myline} & \multicolumn{3}{c}{\myline~FMNIST~\myline} \\
     Method & $M$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ \\
    \midrule
    %\multirow{2}{*}{MNIST}
    MAP & -- & \val{98.22}{0.13} & \val{0.061}{0.004} & \val{0.006}{0.001}
          & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    BNN predictive \cite{immer2021improving} & -- & \val{93.14}{0.05} & \val{0.304}{0.002} & \val{0.111}{0.003}
                                               & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    BNN predictive \cite{ritter2018kfac} & -- & \val{93.03}{0.13} & \val{0.369}{0.003} & \val{0.168}{0.001}
                                           & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    GLM predictive \cite{immer2021improving} & -- & \val{98.40}{0.05} & \val{0.054}{0.002} & \val{0.007}{0.001}
                                               & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    GP predictive \cite{immer2021improving} & $3200$ & \val{98.22}{0.13} & \val{0.058}{0.001} & \val{0.003}{0.000}
                                                     & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\

    GP predictive (NN) & $1024$ & \val{97.70}{0.07} & \val{0.080}{0.002} & \val{0.011}{0.001}
                              & \val{91.47}{0.41} & \val{0.289}{0.009} & \val{0.033}{0.003} \\

    GP predictive & $1024$  & \val{76.28}{1.85} & \val{0.702}{0.029} & \val{0.070}{0.022}
                            & \val{83.98}{0.35} & \val{0.503}{0.012} & \val{0.052}{0.006} \\
    \our (NN) & $1024$ &  \val{98.02}{0.12} & \val{0.066}{0.003} & \val{0.007}{0.001}
                     & \val{91.93}{0.45} & \val{0.267}{0.011} & \val{0.028}{0.003} \\
    \our & $1024$ & \val{98.07}{0.06} & \val{0.063}{0.003} & \val{0.005}{0.001}
                  & \val{91.56}{0.35} & \val{0.253}{0.006} & \val{0.012}{0.001} \\
%    \multirow{2}{*}{FMNIST}
%    & MAP & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    %& BNN predictive \cite{immer2021improving} & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    % & BNN predictive \cite{ritter2018kfac} & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    % & GLM predictive \cite{immer2021improving} & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    %& GP predictive \cite{immer2021improving} $m=3200$ & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\
    % & GP predictive NN $m=1024$ &  \val{0}{0.07} & \val{0}{0.002} & \val{0}{0.001} \\
    % & GP predictive $m=1024$  &  \val{0}{1.85} & \val{0}{0.029} & \val{0}{0.022} \\
    % & \our NN $m=1024$ &  \val{0}{0.0} & \val{0}{0.003} & \val{0}{0.001} \\
    % & \our $m=1024$ &  \val{0}{0.06} & \val{0}{0.003} & \val{0}{0.001} \\
    \bottomrule
  \end{tabular}

@@

Similarly to the UCI experiments, we seek to demonstrate \our on image data that features more complicated data manifolds and model structures.
We consider the MNIST and Fashion-MNIST (FMNIST) classification tasks, and use an MLP and a CNN architecture respectively (matching the setup in \cite{immer2021improving}).
After training, we computed the duals for \our and use the model for prediction (full experiment details in \cref{app:image}).
In \cref{tbl:imagesuper}, we compare to baselines also used in \cite{immer2021improving}.
We report accuracy (ACC), NLPD, and expected calibration error (ECE).
For \our and the GP predictive subset method by \cite{immer2021improving}, we provide results when predicting with GP mean and when using the neural network (NN) and
fix the number of inducing points $M=1024$.
For completeness, we also include the GP predictive result from \cite{immer2021improving} which used more inducing points $M=3200$.
The results in \cref{tbl:imagesuper} agree with the conclusions drawn from the UCI experiments showing that \our is able to more efficiently capture the posterior using a sparse set of points.
They further demonstrate that \our's predictive mean (GP) significantly outperforms the GP predictive subset.
This is an interesting result as we cannot rely on the NN's prediction if we wish to condition on new data, using the dual conditioning from \label{sec:sequential}.
In this setting, we must rely on our GP's predictive mean.
% Other approaches, \cite{immer2021improving} never inten
% The resutls demonstrate that \our's predictive GP mean significantly outperforms the
These results on image data sets motivate experiments beyond the supervised learning setting, which we now detail.

% For \our and the GP predictive subset method by \cite{immer2021improving} we fix the number of points $M=1024$.

%We performed hyperparameter optimization over the prior precision, and found the optimal value to be [insert]. We ran the experiment over $5$ seeds. During the NN training, we used a batch size of $512$ and an Adam optimizer with learning rate $1e-3$.

@@

%% Comments on the results:
From \cref{tbl:cl_table_1}, it is clear that the weight-space regularization methods completely fail compared to the function-space methods on S-MNIST and S-FMNIST, but are comparable on P-MNIST. %, commonly referred as Domain Incremental Learning. (opposed to the Class-Incremental Learning for S-MNIST and S-FMNIST)
Considering only function-space methods, we can see that \our achieves the best results on most data sets, and is particularly effective, when using fewer points per task.
On S-FMNIST, our method obtains close results to DER, which regularizes the model by taking the mean squared error between current and old function outputs without accounting the covariance structure of the loss similar to our L2 ablation. Moreover, their method exploits reservoir sampling \citep{vitter1985random} to obtain a continuously updated set inducing points.
S-FSVI obtains the best on P-MNIST, but with the downside of requiring high computational costs based on performing variational inference, compared to \our. \todo{on PMNIST our method is not really super-fast, so that could be not the best thing to say here} % we could also say that P-MNIST is known to be a bad/ill-posed dataset for CL 
Finally, it is worth noticing as our ablation already provides a very solid baseline consistently over all data sets.

@@

\section{Discussion and conclusion}
\label{sec:conclusion}
%
In this paper, we have introduced \our, a novel approach for representing neural networks in sparse function space, exploiting the dual parameters for an efficient low-rank approximation that accommodates information from the entire data distribution. Our method offers a powerful mechanism for capturing predictive uncertainty, updating the model with new data without retraining, and providing a compact representation suitable for continual learning. These aspects were demonstrated in a wide range of problems, data sets, and learning contexts. We showcased our method's ability to capture uncertainty in UCI classification tasks (\cref{sec:uci}), demonstrated robustness on image data sets (\cref{,sec:image}), established its potential for continual learning (\cref{sec:cl-exp}), and finally, verified its applicability in reinforcement learning scenarios (\cref{sec:rl-exp}).

>>>>

Using the Hessian of the approximate model, we arrive at the Laplace-GGN approximate posterior for $q(\vw)$ \todo{This is the first time q(w) is used, should define it}. 
Therefore, the Laplace-GGN linear model can be used to convert our weight space model to the function space by \todo{To be super clear even when reading this part out of context: could specify that using a GP with this mean and covariance},
\begin{equation} 
\label{eq-laplace-approx-function-space} 
% g(\vx) \sim \GP \left( \mu(\vx), \kappa(\vx, \vx') \right) \quad \text{with} \quad
  \mu(\vx) =  0 \quad \text{and} \quad
  \kappa(\vx, \vx')
  = \frac{1}{\delta} \Jac{\weights^*}{\vx} \, \JacT{\weights^*}{\vx'}, 
\end{equation}
where the kernel is the so-called Neural Tangent Kernel (NTK, \cite{jacot2018neural}). In \cref{eq-laplace-approx-function-space}, we restrict the function output to a single dimension for notational simplicity, yet this formulation can be readily extended to the multi-output scenario. With this kernel function, we can combine it with the data set $\dataset$ to construct the posterior. \todo{what is "it" here?}
%
\citet{khan2019approximate} took a similar approach, but instead of fitting the GP posterior to the actual $\vy$, they rely on a transformation of $\vy$. Similarly, \citet{immer2021improving} obtain the same covariance function but use a different mean because they rely on a first-order approximation of the neural network to form a Bayesian GLM model. Both approaches attempted to adjust the GP posterior mean function to ensure the predictions are from the neural network $f_{\vw^*}(\vx)$. 
Contrarily, our methodology intends to make predictions directly from the derived GP, thus evading such adjustments. Next, we will present the general formulation of the sparse functional representation of our trained neural network based on the non-sparse approximation.

@@

    % THE TABLE CAN BE FORMATTED MORE LIKE THIS
    % \begin{tabular}{l l C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    % \toprule
    % & Method & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & OOD-AUC~$\uparrow$  \\
    % \midrule
    % \multirow{2}{*}{FMNIST}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \midrule
    % \multirow{2}{*}{CIFAR-10}
    % & MAP & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & BNN predictive \cite{todo} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GLM predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & GP predictive & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our (NN) & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % & \our & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} & \val{0.000}{0.000} \\
    % \bottomrule
    % \end{tabular}
  \newcommand{\myline}{\protect\tikz[baseline=-.5ex,line width=.7pt]\protect\draw[draw=gray](0,0)--(7em,0);}
  \begin{tabular}{l c C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw} C{\tblw}}
    \toprule
            &     &  \multicolumn{3}{c}{\myline~MNIST~\myline} & \multicolumn{3}{c}{\myline~FMNIST~\myline} \\
     Method & $M$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ & ACC~$\uparrow$ & NLPD~$\downarrow$ & ECE~$\downarrow$ \\
    \midrule
    %\multirow{2}{*}{MNIST}
    MAP & & \val{98.22}{0.13} & \val{0.061}{0.004} & \val{0.006}{0.001}
          & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    BNN predictive \cite{immer2021improving} & & \val{93.14}{0.05} & \val{0.304}{0.002} & \val{0.111}{0.003}
                                               & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    BNN predictive \cite{ritter2018kfac} & & \val{93.03}{0.13} & \val{0.369}{0.003} & \val{0.168}{0.001}
                                           & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    GLM predictive \cite{immer2021improving} & & \val{98.40}{0.05} & \val{0.054}{0.002} & \val{0.007}{0.001}
                                               & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    GP predictive \cite{immer2021improving} & $3200$ & \val{\textbf{98.22}}{0.13} & \val{\textbf{0.058}}{0.001} & \val{\textbf{0.003}}{0.000}
                                                     & \val{91.36}{0.11} & \val{\textbf{0.25}}{0.004} & \val{\textbf{0.007}}{0.001} \\

    GP predictive (NN) & $1024$ & \val{97.70}{0.07} & \val{0.080}{0.002} & \val{0.011}{0.001}
                              & \val{\textbf{91.47}}{0.41} & \val{0.289}{0.009} & \val{0.033}{0.003} \\

    GP predictive & $1024$  & \val{76.28}{1.85} & \val{0.702}{0.029} & \val{0.070}{0.022}
                            & \val{83.98}{0.35} & \val{0.503}{0.012} & \val{0.052}{0.006} \\
    \our (NN) & $1024$ &  \val{98.02}{0.12} & \val{0.066}{0.003} & \val{0.007}{0.001}
                     & \val{\textbf{91.93}}{0.45} & \val{0.267}{0.011} & \val{0.028}{0.003} \\
    \our & $1024$ & \val{\textbf{98.07}}{0.06} & \val{0.063}{0.003} & \val{0.005}{0.001}
                  & \val{\textbf{91.56}}{0.35} & \val{\textbf{0.253}}{0.006} & \val{0.012}{0.001} \\
%    \multirow{2}{*}{FMNIST}
%    & MAP & \val{91.39}{0.11} & \val{0.258}{0.004} & \val{0.017}{0.001} \\
    %& BNN predictive \cite{immer2021improving} & \val{84.42}{0.12} & \val{0.942}{0.016} & \val{0.411}{0.008} \\
    % & BNN predictive \cite{ritter2018kfac} & \val{91.20}{0.07} & \val{0.265}{0.004} & \val{0.024}{0.002} \\
    % & GLM predictive \cite{immer2021improving} & \val{92.25}{0.10} & \val{0.244}{0.003} & \val{0.012}{0.003} \\
    %& GP predictive \cite{immer2021improving} $m=3200$ & \val{91.36}{0.11} & \val{0.25}{0.004} & \val{0.007}{0.001} \\
    % & GP predictive NN $m=1024$ &  \val{0}{0.07} & \val{0}{0.002} & \val{0}{0.001} \\
    % & GP predictive $m=1024$  &  \val{0}{1.85} & \val{0}{0.029} & \val{0}{0.022} \\
    % & \our NN $m=1024$ &  \val{0}{0.0} & \val{0}{0.003} & \val{0}{0.001} \\
    % & \our $m=1024$ &  \val{0}{0.06} & \val{0}{0.003} & \val{0}{0.001} \\
    \bottomrule
  \end{tabular}

@@

Similarly to the UCI experiments, we seek to demonstrate \our on image data that features more complicated data manifolds and model structures. We consider the MNIST and Fashion-MNIST (FMNIST) classification tasks, and use an MLP and a CNN architecture respectively (matching the setup in \cite{immer2021improving}). After training, we computed the duals for \our and use the model for prediction (full experiment details in \cref{app:image}). In \cref{tbl:imagesuper}, we compare to baselines also used in \cite{immer2021improving}. For \our and the GP predictive subset method by \cite{immer2021improving} we fix the number of points $M=1024$. For completeness, we include the GP predictive result from \cite{immer2021improving} which used more inducing points $M=3200$. We report accuracy (ACC), NLPD, and expected calibration error (ECE). The results in \cref{tbl:imagesuper} agree with the conclusions drawn from the UCI experiments showing that \our is able to more efficiently capture the posterior using a sparse set of points. These results on image data sets motivate experiments beyond the supervised learning setting, which we now detail.
%We performed hyperparameter optimization over the prior precision, and found the optimal value to be [insert]. We ran the experiment over $5$ seeds. During the NN training, we used a batch size of $512$ and an Adam optimizer with learning rate $1e-3$.

@@

%% Comments on the results:
From \cref{tbl:cl_table_1}, it is clear that the weight-space regularization methods completely fail compared to the function-space methods on S-MNIST and S-FMNIST, but are comparable on P-MNIST. %, commonly referred as Domain Incremental Learning. (opposed to the Class-Incremental Learning for S-MNIST and S-FMNIST)
Considering only function-space methods, we can see that \our achieves the best results on most datasets, and is particularly effective, when using fewer points per task.
On S-FMNIST, our method obtains close results to DER, which regularizes the model by taking the mean squared error between current and old function outputs without accounting the covariance structure of the loss similar to our L2 ablation. Moreover, their method exploits reservoir sampling \citep{vitter1985random} to obtain a continuously updated set inducing points.
S-FSVI obtains the best on P-MNIST, but with the downside of requiring high computational costs based on performing variational inference, compared to \our. \todo{on PMNIST our method is not really super-fast, so that could be not the best thing to say here} % we could also say that P-MNIST is known to be a bad/ill-posed dataset for CL 
Finally, it is worth noticing as our ablation already provides a very solid baseline consistently over all datasets.

@@

\section{Discussion and conclusion}
\label{sec:conclusion}
%
In this paper, we have introduced \our, a novel approach for representing neural networks in sparse function space, exploiting the dual parameters for an efficient low-rank approximation that accommodates information from the entire data distribution. Our method offers a powerful mechanism for capturing predictive uncertainty, updating the model with new data without retraining, and providing a compact representation suitable for continual learning. These aspects were demonstrated in a wide range of problems, datasets, and learning contexts. We showcased our method's ability to capture uncertainty in UCI classification tasks (\cref{sec:uci}), demonstrated robustness on image data sets (\cref{,sec:image}), established its potential for continual learning (\cref{sec:cl-exp}), and finally, verified its applicability in reinforcement learning scenarios (\cref{sec:rl-exp}).
